D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-09 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
Here are some small edits for the section on hardware.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Dependencies between machine instructions can be understood in terms
of the flow of information within a CPU.
Each instruction has a set of inputs and is the source of a set of outputs,
some of which flow to the inputs for later instructions.
A few examples:
\begin{itemize}
\item	An \co{add} instruction would typically have two inputs
	(the register values to be added together) and two outputs
	(the sum to be stored in a general-purpose register and some
	flag bits---e.g., Zero, Carry, and Overflow---to be stored in
	a status register).
\item	A conditional branch instruction would have only one input:
	the flag bits it tests in the status register.
	A computed branch instruction would as input the register
	containing the address to branch to.
	For both, the output is the new address to be written to the
	instruction-pointer register.
\item	A memory-load instruction's input is the address to load from,
	and its output is the value obtained by the load, to be stored
	in a register.
\item	A memory-store instruction's inputs are the value to store and
	the address at which to store it; there are no outputs.
\end{itemize}
Of course, some of an instruction's inputs may be immediate constants
taken from the instruction itself rather than outputs from some
earlier instruction.
Note: Hardware dependency analysis considers only the information
flowing \emph{within} a CPU, not the information flowing between the
CPU and memory, which is handled separately.

Using this scheme, we say that an instruction $J$ is dependent on
another instruction $I$ when any of $I$'s outputs flow into $J$'s inputs,
perhaps indirectly via some intermediate instructions.
Tracing back the flow of information between instructions, you can
see that any hardware dependency must ultimately emanate from a load
instruction or a thread's initial state (or possibly some sort of
input instruction, but we will not consider that here).
These are the only sources of truly new information in a thread.

@@

A CPU may start executing an instruction speculatively, but at some
point it must \emph{commit} to a decision either to definitely execute the
instruction with some well-defined collection of inputs and outputs
or else to abandon it.
Thanks to the law of cause and~effect, a CPU is not able to commit an
instruction until all of its inputs' sources have committed.
(Or at least, all that can affect the action and outputs of the
instruction.
For instance, although no existing architectures do this, there is nothing
in principle to prevent a CPU from committing a \co{multiply}
instruction as soon as either one of its inputs has committed to the
value zero.)
The reason is obvious: An input is subject to change at any time until
its source commits, the outputs aren't determined until the inputs
are, and the instruction can't commit until its outputs are fully
determined.

This means that, with a few possible exceptions such as the
multiply-by-zero example, an instruction can't commit until all the
instructions it depends on have committed.
And even taking the exceptions into account, if a change in the
outputs of instruction $I$ would lead to a change in the inputs to
instruction $J$ then $I$ must commit before $J$.
Thus dependencies force instructions to commit in order, even on
weakly ordered architectures.

However, dependencies aren't the only mechanism capable of forcing
ordering in hardware executions.
The simplest of these is a load reading from a store in another
thread, i.e., the rfe relation.
As mentioned in section~\ref{sec:Simple OOTA Cycle} above, a CPU does
not make the value of a store instruction available for other CPUs to
load until the store commits or later.
And after this happens, it takes some time for the information about
the store to travel from that CPU to others, owing to the inescapable
facts that processors have nonzero size and information cannot travel
faster than the speed of light.
Since the load instruction cannot commit until its output (the value it
reads) is fully determined, it must commit after the store it reads
from.

A similarly straightforward mechanism for ordering is conditional or
computed branches.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to;
until then the CPU cannot know whether the instruction should
be executed at all.
Therefore instructions following a conditional or computed branch
instruction in an execution must commit after anything the branch
depends on.
(Again, there could theoretically be an exception for a branch that
conditionally jumps to the immediately following instruction.
We can ignore this issue by treating such branches as no-ops.)

Yet another mechanism for ordering crops up when a load reads from a
store in the same thread (the rfi relation).
In this situation the load cannot commit before the sources for the
store's inputs.
Until the source for the store's value commits, the value of the load
isn't fully determined and so the load cannot commit.
And in general, until the source for store's address commits, it's not
fully determined whether the store and load are for the same address.
This argument doesn't hold if the store and load have the \emph{same}
source for their addresses, but another factor is also at play.
Until the address is fully determined, there is a possibility that it
will turn out to be invalid (a NULL pointer, for example), causing the
store instruction to encounter an addressing exception.
If this should happen the load might not execute at all, and therefore
it can't commit until the address's source has committed.

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as partially volatile.
\begin{quote}
Let $W$ be a store which, in some abstract execution $A$, semantically
depends on loads $\{R_0$, \ldots,~$R_n\}$, and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted and the hardware
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}
It's quite straightforward to show that for any implementation in which
semantic dependencies satisfy the Fundamental Property, no OOTA
abstract execution can be realized.

Indeed, suppose that abstract execution $A$ is realized by hardware
execution $H$ and $A$ has an OOTA cycle.
This means there are stores $W_i$ semantically depending on loads
$\{R_{i,j}\}$ where each of the loads reads from one of the $W_i$
stores in a different thread.
Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
map to, if they aren't omitted.
Assuming that at least one of the stores is not omitted,
one of the $W_i'$, call it $W_{i_0}'$, must commit first.
By the Fundamental Property, one of the loads that $W_{i_0}'$ depends on,
call it $R_{i_0,j_0}'$, is not omitted and is ordered before $W_{i_0}'$.
But now we have a contradiction:
\begin{itemize}
\item	$W_{i_0}'$ commits after $R_{i_0,j_0}'$;
\item	$R_{i_0,j_0}'$ commits after the store $W_k'$ it reads from;
\item	$W_k'$ commits no earlier than $W_{i_0}'$.
\end{itemize}

If all the stores in the OOTA cycle are omitted then all the reads
must be omitted as well.
This amounts to a situation where the entire cycle has been optimized
away by the compiler.

>>>>

Dependencies between machine instructions can be understood in terms
of the flow of information within a CPU.
Each instruction has a set of inputs and is the source of a set of outputs,
some of which flow to the inputs for later instructions.
A few examples:
\begin{itemize}
\item	An \co{add} instruction would typically have two inputs
	(the register values to be added together) and two outputs
	(the sum to be stored in a general-purpose register and some
	flag bits---e.g., Zero, Carry, and Overflow---to be stored in
	a status register).
\item	A conditional branch instruction would have only one input:
	the flag bits it tests in the status register.
	A computed branch instruction would have as input the register
	containing the address to branch to.
	For both, the output is the new address to be written to the
	instruction-pointer register.
\item	A memory-load instruction's input is the address to load from,
	and its output is the value obtained by the load, to be stored
	in a register.
\item	A memory-store instruction's inputs are the value to store and
	the address at which to store it; there are no outputs.
\end{itemize}
Of course, some of an instruction's inputs may be immediate constants
taken from the instruction itself rather than outputs from some
earlier instruction.
Note: Hardware dependency analysis considers only the information
flowing \emph{within} a CPU, not information flowing between the
CPU and memory, which is handled separately.

Using this scheme, we say that an instruction $J$ is dependent on
another instruction $I$ when any of $I$'s outputs flow into $J$'s inputs,
perhaps indirectly via some intermediate instructions.
Tracing back the flow of information between instructions, you can
see that any hardware dependency must ultimately emanate from a load
instruction or the thread's initial register values (or possibly some sort of
input instruction, but we will not consider that here).
These are the only sources of truly new information in a thread.

@@

A CPU may start executing an instruction speculatively, but at some
point it must \emph{commit} to a decision either to definitely execute the
instruction's action with some collection of well defined inputs and outputs
or else to abandon it.
Thanks to the law of cause and~effect, a CPU is not able to commit an
instruction until all of its inputs' sources have committed.
(Or at least, all that can affect the action and outputs of the
instruction.
For instance, although no existing architectures do this, there is nothing
in principle to prevent a CPU from committing a \co{multiply}
instruction as soon as either one of its inputs' sources has committed to the
value zero.)
The reason is obvious: An input is subject to change at any time until
its source commits,
the action and outputs aren't fully determined until the inputs
are, and the instruction can't commit until its action and outputs are fully
determined.

This means that, with a few possible exceptions such as the
multiply-by-zero example, an instruction can't commit until all the
instructions it depends on have committed.
And even taking the exceptions into account, if a change in the
outputs of instruction $I$ would lead to a change in the action or outputs of
instruction $J$ then $I$ must commit before $J$.
Thus dependencies force instructions to commit in order, even on
weakly ordered architectures.

However, dependencies aren't the only mechanism capable of forcing
ordering in hardware executions.
The simplest of these is a load reading from a store in another
thread, i.e., the rfe relation.
As mentioned in section~\ref{sec:Simple OOTA Cycle} above, a CPU does
not make the value of a store instruction available for other CPUs to
load until the store commits or some time later.
And after this happens, it takes some time for the information about
the store to travel from that CPU to others, owing to the inescapable
facts that processors have nonzero size and information cannot travel
faster than the speed of light.
Since the load instruction cannot commit until its output (the value it
reads) is fully determined, it must commit after the store it reads
from.

A similarly straightforward mechanism for ordering is conditional or
computed branches.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to;
until then the CPU cannot know whether the later instruction should
be executed at all.
Therefore instructions following a conditional or computed branch
instruction in an execution must commit after anything the branch
depends on.
(Again, there could theoretically be an exception for a branch that
conditionally jumps to the immediately following instruction.
We can ignore this issue by treating such branches as no-ops.)

Yet another mechanism for ordering crops up when a load reads from a
store in the same thread (the rfi relation).
In this situation the load cannot commit before the sources for the
store's inputs.
Until the source for the store's value commits, the value of the load
isn't fully determined and so the load cannot commit.
And in general, until the source for store's address commits, it's not
fully determined whether the store and load will access the same address.
This argument doesn't hold if the store and load have the \emph{same}
source for their addresses, but another factor is also at play.
Until the address is fully determined, there is a possibility that it
will turn out to be invalid (a NULL pointer, for example), causing the
store instruction to generate an addressing exception.
If this should happen the load might not execute at all, and therefore
it can't commit until the address's source has committed.

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as partially volatile.
\begin{quote}
Let $W$ be a store which, in some abstract execution $A$, semantically
depends on loads $\{R_0$, \ldots,~$R_n\}$, and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted and the hardware
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}
It's quite straightforward to show that for any implementation in which
semantic dependencies satisfy the Fundamental Property, no OOTA cycle
has a nontrivial realization.

Indeed, suppose that abstract execution $A$ is realized by hardware
execution $H$ and $A$ has an OOTA cycle.
This means there are stores $W_i$ in $A$, semantically depending on loads
$\{R_{i,j}\}$ where each of the loads reads from one of the $W_k$
stores in a different thread.
Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
map to in $H$, if they aren't omitted.
Assuming that at least one of the stores is not omitted,
one of the $W_i'$ instructions, let's say $W_0'$, must commit first.
By the Fundamental Property, one of the loads that $W_0$ depends on,
let's say $R_{0,0}$, is not omitted and $R_{0,0}'$ is ordered before $W_0'$.
But now we have a contradiction:
\begin{itemize}
\item	$W_0'$ commits after $R_{0,0}'$;
\item	$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
\item	$W_k'$ commits no earlier than $W_0'$.
\end{itemize}

If all the stores in the OOTA cycle are omitted then all the reads
must be omitted as well.
This amounts to a situation where the entire cycle has been optimized
out of existence by the compiler.
Although we can are unable to prove it, we conjecture that in this
situation there must be another abstract execution which has the
same observable effects as $A$ and is also realized by $H$, but in
which the OOTA cycle does not occur.
Thus there would be no way to tell, merely by observing the effects of
$A$, whether there was an OOTA cycle or not.
For this reason we declare realizations of OOTA cycles in which
all the accesses are omitted to be trivial.
